A Hitchhiker's Guide to Reproducible Research in R

### 2021-07-07 manuscript.Rmd
cite further materials for Git closes #133
explain how to extend Make

We recommend using **Git** for version control, because of its widespread adoption in the R community.
Git is built around snapshots that represent the entire project state at a given point in time.
Those snapshots are called "commits".
Each file that was changed has to be added to a commit, and each commit requires a commit message that succinctly describes the changes made.
Each commit "knows" its ancestor, and they form a timeline together.
The conglomerate of commits (i.e. the version-controlled project) is called a repository.
In Git, specific snapshots of a repository can be tagged, such that one can clearly label a given project status as the one that created a given preregistration, a given preprint, or the final authors' version accepted at a journal.
Git has additional features beyond basic version control, such as branches to facilitate collaboration.
To collaborate via Git,<!--dangling modify, cant help it--> the repository has to be uploaded somewhere.
We recommend GitHub to host Git repositories because of its popularity among R users.
GitHub has many tools that ease project management and collaboration, and these tools provide much value in our everyday work, but they are not central to archiving reproducibility.

@@

```{bash, eval=FALSE}
manuscript.pdf: manuscript.Rmd simulation_results.csv 
  Rscript -e 'rmarkdown::render("manuscript.Rmd")'

simulation_results.csv: simulate.R
  Rscript -e 'source("simulate.R")'
```

There are two targets, the final rendered report (`manuscript.pdf`, l.1) and the simulation results (`simulation_results.csv`, l.4).
Each target is followed by a colon and a list of requirements.
If a requirement is newer than the target, the recipe will be executed to rebuild the target.
If a requirement does not exist, Make uses a recipe to build the requirement first before building the target.
Here, if one were to build the final `manuscript.pdf` by rendering the R Markdown with the command shown in l.2, Make would make sure that the file `simulation_results.csv` exists; if not, it would find that this is also a target and issue the command in l.5 to run the simulation first before rendering the manuscript.
This way, we ensure that the simulation is always run before the manuscript is built and that the manuscript is rebuilt if the simulation code was changed.

A version controlled dynamic document with dependency tracking still relies on external software.
Troubleshooting issues specific to a particular programming language or dependent tool typically requires considerable expertise and hinder reproducibility.

@@

One solution to the problem of external software dependencies is to provide not only the analysis script but all dependent software packages and system libraries via **software management**.
One comprehensive approach to software management is containerization.
"By packaging the key elements of the computational  environment  needed  to  run  the  desired  software, [...] they  make the software much easier to use, and the results easier to reproduce [@silverSoftwareSimplified2017].
Instead of leaving it to the user which software should be installed how on their computer, **Docker** provides all software used in an analysis, including the operating system itself.
The operating system level is important because some functionality may be passed to software layers beneath the programming language, such as calls to random number generators, linear algebra libraries and such.
Docker does this without interfering with the already installed software by using a virtual software environment independent of the host software environment.
Such a snapshot of the software stack is called an "image".
Packaging all needed software in such an image requires considerable amounts of storage space.
Two major strategies are helping to keep the storage requirements reasonable.
First, there is a community that maintains images for particular purposes.
For example, there are ready-made images that only include what is necessary for R, based on Ubuntu (a Linux operating system) containers [@boettigerIntroductionRockerDocker2017].
Users can then install whatever they need in addition to what is provided by these pre-compiled images.
`r ifelse(is.na(container_size), "", glue::glue("The image that was used for this article uses {round(container_size, 2)}Gb of disc space."))`
Our image includes Ubuntu, R, R Studio, LaTeX as well as a variety of R packages like the tidyverse [@tidyverse], amounting to `r installed.packages() %>% as.data.frame() %>% filter(is.na(Priority)) %>% nrow()` packages.
Second, instead of saving a binary image of the software, one can also save a container recipe, a so-called  `Dockerfile`.
This recipe is a textual description of all commands that need to be executed to recreate the image.
Such files are tiny (the [Dockerfile](https://github.com/aaronpeikert/repro-tutorial/blob/main/Dockerfile) for this project is `r round(as.numeric(fs::file_size("Dockerfile"))/1024^1, 2)`Kb big) and are an sufficently describe how to recreate a given container image.
However, they rely on the assumption that all software repositories will remain accessible and will continue to make available historic software versions.
For archiving, we therefore recommend that a full image is always kept (see Section on [Archiving Docker]).
A more comprehensive overview over how to employ containarization in research projects is given by @wiebelsLeveragingContainersReproducible2021.

To summarize, the workflow by @Peikert2019 requires four components (see Figure \@ref(fig:schematic)), dynamic document generation (using R Markdown), version control (using Git), internal dependency management (using Make), containerization (using Docker).
While R Markdown and Git are well integrated into the R environment through R Studio, Make and Docker require a level of expertise that is often beyond the training of scholars outside studies of information technology, which represents a considerable hurdle in the acceptance and implementation of the workflow.
To remove this hurdle, we have developed a new R package `repro` that supports scholars in setting up, maintaining, and reproducing research projects in R.
In the remainder, we will walk you through a complete research project, from sketching the very first idea over preregistration, data analysis, and submission of a preprint while focusing on how the `repro`-package enables users to maintain reproducibility of all steps.

@@

They detect the users' system state and explain what this means to the user.
Sometimes they ask the user to do something, e.g. the following happens if you were a windows user who does not have installed Git:

@@

This type of markup for formatting, instead of the more visual approach of e.g. Microsoft Word ©, is compatible with many output formats and is more easily version controlled.
However, researchers might find it easier to activate the "Visual Editor" of R Studio (Ctrl + Shift + F4 or click on the icon that resembles drawing materials or a compass in the upper right corner of the R Markdown document) which features more graphical elements like a traditional word processor with still creates an R Markdown and has all of its benefits.
The Visual Editor has the additional benefit of promoting best practices, e.g. each sentence on a newline for better version control or automated citations, cross references and more.

@@

The file `.repro/Dockerfile_manual` is by default empty and is supposed to allow dependencies outside of R, like system libraries or external software.
Remember, the software environment of the Docker image is completely separate from software installed on your computer.
This separation is excellent for reproducibility, but requires you to install software on an operating system that may not be familiar to you.
The images supplied by [@boettigerIntroductionRockerDocker2017] are based on Ubuntu, and hence software has to be installed in a way that is compatible with Ubuntu.
The most convenient way to install software on Ubuntu is through its package manager `apt`.
This snipped installs python, added to `.repro/Dockerfile_manual` you get an image that has python available:

@@

The the last Docker, `.repro/Dockerfile_packages`, should not be manually changed.
`repro::automate()` gathers all packages from all `.Rmd` files and figures out if they should be installed from CRAN or GitHub.
`repro::automate()` also simply combines the three Dockerfiles and saves the result into the main `Dockerfile` at the top level of the project.
The seperation into several Dockerfiles is only to provide a clear seperation between automatically generated parts and manually altered parts of the Dockerfile.
However, Docker requires in the end the one Dockerfile that `repro::automate()` creates from them.

@@

PACs come with a challange though, in contrast to a reproducible script, that should work on the same data, a PAC must function on different data.
We call this property "reusability".
This property is based on reproducibility, but requires the researcher to more carefully write the software [@lanerganSoftwareEngineeringReusable1989] so it is build-for-reuse [@al-badareenReusableSoftwareComponents2010].
The reproducible workflow we present here is heavily automated and hence promotes reusibility.
Increased automation is more and more recognized as a means to improve the research process [@rouderMinimizingMistakesPsychological2019] and therefore this workflow fits well together with other innovations employ automation, like machine readible hypothethis tests [@lakensImprovingTransparencyFalsifiability2021] or automated data documentation [@arslanHowAutomaticallyDocument2019].
Automated and reusable research projects promises a wide range of applications, among them PAC [possibly submitted as a registered report @nosekRegisteredReports2014; @chambersWhatNextRegistered2019], Direct Replication[@simonsValueDirectReplication2014], fully automated living metanalysis [@elliottLivingSystematicReviews2014], Executable Research Articles [@elifesciencespublicationsELifeLaunchesExecutable2020], and other innovations like the live analysis of born open data [@rouderWhatWhyHow2016; @kekecsRaisingValueResearch2019].

>>>>

We recommend using **Git** for version control, because of its widespread adoption in the R community.
Git is built around snapshots that represent the entire project state at a given point in time.
Those snapshots are called "commits".
Each file that was changed has to be added to a commit, and each commit requires a commit message that succinctly describes the changes made.
Each commit "knows" its ancestor, and they form a timeline together.
The conglomerate of commits (i.e. the version-controlled project) is called a repository.
In Git, specific snapshots of a repository can be tagged, such that one can clearly label a given project status as the one that created a given preregistration, a given preprint, or the final authors' version accepted at a journal.
Git has additional features beyond basic version control, such as branches to facilitate collaboration.
@vuorreCuratingResearchAssets2018 provide a more extensive treatment how Git functions and how to use Git.
@bryanExcuseMeYou2018 provides additional information on how to track R Markdown documents.
To collaborate via Git,<!--dangling modify, cant help it--> the repository has to be uploaded somewhere.
We recommend GitHub to host Git repositories because of its popularity among R users.
GitHub has many tools that ease project management and collaboration, and these tools provide much value in our everyday work, but they are not central to archiving reproducibility.

@@

```{bash, eval=FALSE}
manuscript.pdf: manuscript.Rmd simulated_data.csv
  Rscript -e 'rmarkdown::render("manuscript.Rmd")'

simulated_data.csv: simulate.R
  Rscript -e 'source("simulate.R")'
```

There are two targets, the final rendered report (`manuscript.pdf`, l.1) and the simulation results (`simulation_results.csv`, l.4).
Each target is followed by a colon and a list of requirements.
If a requirement is newer than the target, the recipe will be executed to rebuild the target.
If a requirement does not exist, Make uses a recipe to build the requirement first before building the target.
Here, if one were to build the final `manuscript.pdf` by rendering the R Markdown with the command shown in l.2, Make would make sure that the file `simulation_results.csv` exists; if not, it would find that this is also a target and issue the command in l.5 to run the simulation first before rendering the manuscript.
This way, we ensure that the simulation is always run before the manuscript is built and that the manuscript is rebuilt if the simulation code was changed.
Make therefore offers a standardized process to reproduce projects, regardless of the complexity or configuration of the project.

@@

A version controlled dynamic document with dependency tracking still relies on external software.
Troubleshooting issues specific to a particular programming language or dependent tool typically requires considerable expertise and hinder reproducibility.
One solution to the problem of external software dependencies is to provide not only the analysis script but all dependent software packages and system libraries via **software management**.
One comprehensive approach to software management is containerization.
"By packaging the key elements of the computational  environment  needed  to  run  the  desired  software, [...] they  make the software much easier to use, and the results easier to reproduce [@silverSoftwareSimplified2017].
Instead of leaving it to the user which software should be installed how on their computer, **Docker** provides all software used in an analysis, including the operating system itself.
The operating system level is important because some functionality may be passed to software layers beneath the programming language, such as calls to random number generators, linear algebra libraries and such.
Docker does this without interfering with the already installed software by using a virtual software environment independent of the host software environment.
Researchers therefore have the opportunity to use the exact same software that was used by the original authors.
Such a snapshot of the software stack is called an "image".
Packaging all needed software in such an image requires considerable amounts of storage space.
Two major strategies are helping to keep the storage requirements reasonable.
First, there is a community that maintains images for particular purposes.
For example, there are ready-made images that only include what is necessary for R, based on Ubuntu (a Linux operating system) containers [@boettigerIntroductionRockerDocker2017].
Users can then install whatever they need in addition to what is provided by these pre-compiled images.
`r ifelse(is.na(container_size), "", glue::glue("The image that was used for this article uses {round(container_size, 2)}Gb of disc space."))`
Our image includes Ubuntu, R, R Studio, LaTeX as well as a variety of R packages like the tidyverse [@tidyverse], amounting to `r installed.packages() %>% as.data.frame() %>% filter(is.na(Priority)) %>% nrow()` packages.
Second, instead of saving a binary image of the software, one can also save a container recipe, a so-called  `Dockerfile`.
This recipe is a textual description of all commands that need to be executed to recreate the image.
Such files are tiny (the [Dockerfile](https://github.com/aaronpeikert/repro-tutorial/blob/main/Dockerfile) for this project is `r round(as.numeric(fs::file_size("Dockerfile"))/1024^1, 2)`Kb big) and sufficiently describe how to recreate a given container image.
However, they rely on the assumption that all software repositories will remain accessible and will continue to make available historic software versions.
For archiving, we therefore recommend that a full image is always kept.
A more comprehensive overview over how to employ containarization in research projects is given by @wiebelsLeveragingContainersReproducible2021.

To summarize, the workflow by @Peikert2019 requires four components (see Figure \@ref(fig:schematic)), dynamic document generation (using R Markdown), version control (using Git), internal dependency management (using Make), containerization (using Docker).
While R Markdown and Git are well integrated into the R environment through R Studio, Make and Docker require a level of expertise that is often beyond the training of scholars outside studies of information technology, which represents a considerable hurdle in the acceptance and implementation of the workflow.
To remove this hurdle, we have developed a new R package `repro` that supports scholars in setting up, maintaining, and reproducing research projects in R.
In the remainder, we will walk you through a the creation of a reproducible research project aided by the package `repro`.

@@

They detect the users' system state and explain what this means to the user.
Sometimes they ask the user to do something, e.g. the following happens if you were a Windows user who does not have installed Git:

@@

This type of markup for formatting, instead of the more visual approach of e.g. Microsoft Word ©, is compatible with many output formats and is more easily version controlled.
However, researchers might find it easier to activate the "Visual Editor" of R Studio (Ctrl + Shift + F4 or click on the icon that resembles drawing materials or a compass in the upper right corner of the R Markdown document) which features more graphical elements like a traditional word processor but still creates an R Markdown and has all of its benefits.
The Visual Editor has the additional benefit of promoting best practices, e.g. each sentence on a newline for better version control or automated citations, cross references and more.

@@

The file `.repro/Dockerfile_manual` is by default empty and is supposed to allow dependencies outside of R, like system libraries or external software.
Remember, the software environment of the Docker image is completely separate from software installed on your computer.
This separation is excellent for reproducibility, but requires you to install software on an operating system that may not be familiar to you.
The images supplied by [@boettigerIntroductionRockerDocker2017] are based on Ubuntu, and hence software has to be installed in a way that is compatible with Ubuntu.
The most convenient way to install software on Ubuntu is through its package manager `apt`.
The following snipped installs Python.
If added to `.repro/Dockerfile_manual` the Docker image will have Python installed.

@@

The last Dockerfile, `.repro/Dockerfile_packages`, should not be manually changed.
`repro::automate()` gathers all packages from all `.Rmd` files and figures out if they should be installed from CRAN or GitHub.

The seperation into several Dockerfiles is only to provide a clear seperation between automatically generated parts and manually altered parts of the Dockerfile.
However, Docker requires single Dockerfile so `repro::automate()` simply combines the three Dockerfiles and saves the result into the main `Dockerfile` at the top level of the

Users of `repro` can not only build arbitrarily complex software environments they can also implement complex file dependencies.
The standard `repro` metadata only makes sure that all dependencies are availible but does not allow you to specify a recipe for them in the metadata.
If you can formulate the creation of dependencies in terms of computational steps you should include these in the Makefile.
The Makefile that `repro` creates is only an template and you are free to change it.
However, you want to make sure to not remove the following two lines:

```bash
include .repro/Makefile_Rmds
include .repro/Makefile_Docker
```

The file `.repro/Makefile_Rmds` contains the automatically generated targets from `repro::automate()` for the R Markdown files.
This file should not be altered manually.
If you want to adept a target than simply provide an alternative target in the main `Makefile`.
Targets in the main `Makefile` take precedent.

The file `.repro/Makefile_Docker` does again contain a rather complicated template which you could but seldomly should alter.
This Makefile coordinates the interplay between Make and Docker and contains targets for building (with `make docker`) and saving (with `make save-docker`) the Docker image.
Additionally it provides facilities to execute commands within the container.
If you write computational recipe for a target it will be by default evaluated using the locally installed software.
To instead evaluate commends inside the Docker image you should wrap them in `$(RUN1) command $(RUN2)`, like in this example:

```bash
simulated_data.csv: R/simulate.R
  $(RUN1) Rscript -e 'source("R/simulate.R")' $(RUN1)
```

If users use

```bash
make simulated_data.csv
```

this is simply translated to:

```{r, echo=FALSE, results='asis'}
cat("```bash",
  stringi::stri_wrap(system2("make", " --dry-run simulated_data.csv", stdout = TRUE)),
  "```", sep = "\n")
```

But if users use

```bash
make DOCKER=TRUE simulated_data.csv
```

it is evaluated within the Docker container:

```{r, echo=FALSE, results='asis'}
cat("```bash",
  stringi::stri_wrap(system2("make", " --dry-run DOCKER=TRUE simulated_data.csv", stdout = TRUE)),
  "```", sep = "\n")
```

@@

PACs come with a challange though, in contrast to a reproducible script, that should work on the same data, a PAC must function on different data.
We call this property "reusability".
This property is based on reproducibility, but requires the researcher to more carefully write the software [@lanerganSoftwareEngineeringReusable1989] so it is build-for-reuse [@al-badareenReusableSoftwareComponents2010].
The reproducible workflow we present here is heavily automated and hence promotes reusibility.
Increased automation is more and more recognized as a means to improve the research process [@rouderMinimizingMistakesPsychological2019] and therefore this workflow fits well together with other innovations that employ automation, like machine readible hypothethis tests [@lakensImprovingTransparencyFalsifiability2021] or automated data documentation [@arslanHowAutomaticallyDocument2019].
Automated and reusable research projects promises a wide range of applications, among them PAC [possibly submitted as a registered report @nosekRegisteredReports2014; @chambersWhatNextRegistered2019], Direct Replication[@simonsValueDirectReplication2014], fully automated living metanalysis [@elliottLivingSystematicReviews2014], Executable Research Articles [@elifesciencespublicationsELifeLaunchesExecutable2020], and other innovations like the live analysis of born open data [@rouderWhatWhyHow2016; @kekecsRaisingValueResearch2019].

Therefore reproducibility facilitates traditional good scientific practices and provides the fundament for promising innovations.
